---
title: "In-class Exercise 4"
date: "9 December 2023"
date-modified: "last-modified" #allows for updating to the latest date
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# In-class Exercise 4: Geospatial Data Science with R

## 1 Overview

In this in-class exercise, the following tasks are performed:

1.  Perform geocoding using data downloaded from Data.gov.sg; and

2.  Calibrate Geographically Weighted Poisson Regression.

## 2 Getting Started

The following packages are loaded into the R environment:

```{r}
pacman::p_load(tidyverse, sf, httr, tmap)
```

## 3 Geocoding using SLA API

Address geocoding, or simply geocoding, is the process of taking an aspatial description of a location, such as an address or postcode, and returning geographic coordinate, frequently

```{r}
url = "https://www.onemap.gov.sg/api/common/elastic/search"

csv = read_csv("data/aspatial/Generalinformationofschools.csv")
postcodes = csv$`postal_code`

found = data.frame()
not_found = data.frame()

for(postcode in postcodes){
  query = list('searchVal' = postcode, 'returnGeom' = 'Y', 'getAddrDetails' = 'Y', 'pageNum' = '1')
  res = GET(url, query = query)
  if((content(res)$found)!=0){
  found = rbind(found, data.frame(content(res))[4:13])
  } else{
  not_found = data.frame(postcode)
  }
}
```

xxx

```{r}
merged = merge(csv, found, by.x = 'postal_code', by.y = 'results.POSTAL', all = TRUE)
write.csv(merged, file = 'data/aspatial/schools.csv')
write.csv(not_found, file = 'data/aspatial/not_found.csv')
```

## 4 Converting Aspatial Data into simple feature tibble data.frame

### 4.1 Importing and Tidying *schools* Data

In this sub-section, the schools.csv is imported into the R environment and only the necessary fields are selected and renamed.

```{r}
schools = read_csv('data/aspatial/schools.csv') %>%
  rename(latitude = "results.LATITUDE", 
         longitude = "results.LONGITUDE") %>%
  select(postal_code, school_name, latitude, longitude)
```

### 4.2 Converting Aspatial Data into sf tibble data.frame

Next, the aspatial data is converted into a simple feature tibble data.frame, *schools_sf*.

```{r}
schools = schools %>%
  mutate(latitude = ifelse(school_name == "ZHENGHUA SECONDARY SCHOOL", 1.389279, latitude)) %>%
  mutate(longitude = ifelse(school_name == "ZHENGHUA SECONDARY SCHOOL", 103.7651, longitude))
```

```{r}
schools_sf = st_as_sf(schools,
                      coords = c("longitude", "latitude"),
                      crs = 4326) %>%
  st_transform(crs = 3414)
```

### 4.3 Plotting A Point Simple Feature Layer

```{r}
tmap_mode("view")
tm_shape(schools_sf) +
  tm_dots() +
  tm_view(set.zoom.limits = c(11, 14))
tmap_mode("plot")
```

## 5 Preparing Data

xxx

```{r}
mpsz = read_rds("data/rds/mpsz.rds")
```

xxx

```{r}
mpsz$`SCHOOL_COUNT` = lengths(
  st_intersects(mpsz, schools_sf)
)
```

The summary statistics of the derived variable shows that the minimum value is 0, maximum value is 12. The data is skewed as the median and 1Q values are also 0 and the mean value is 1.054.

```{r}
summary(mpsz$SCHOOL_COUNT)
```

xxx

```{r}
biz_sf = st_read(dsn = "data/geospatial",
                 layer = "business")
```

xxx

```{r}
tmap_options(check.and.fix = TRUE)
tm_shape(mpsz) + 
  tm_polygons() +
  tm_shape(biz_sf) +
  tm_dots()
```

## 6 Data Integration and Wrangling

xxx

```{r}
# flow_data = flow_data %>%
#  left_join(mpsz_tidy,
#            by = c("DESTIN_SZ" = "SUBZONE_C"))
```

6.1 Checking for Variables with Zero Values

xxx

```{r}
# summary(flow_data)
```

xxx

```{r}

```

------------------------------------------------------------------------

2 Getting Started

```{r}
pacman::p_load(tmap, sf, performance,
               ggpubr, tidyverse)
```

3 Data

```{r}
flow_data = read_rds("data/rds/flow_data_tidy.rds")
```

```{r}
glimpse(flow_data)
```

xxx

```{r}
flow_data$FlowNoIntra = ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0, flow_data$MORNING_PEAK)
flow_data$offset = ifelse(
  flow_data$ORIGIN_SZ == flow_data$DESTIN_SZ,
  0.000001, 1)
```

```{r}
inter_zonal_flow = flow_data %>%
  filter(FlowNoIntra > 0)
```

```{r}
inter_zonal_flow = inter_zonal_flow %>%
  rename(TRIPS = MORNING_PEAK,
         DIST = dist,
         BIZ_COUNT = RETAIL_COUNT)
```

### Unconstrained Spatial Interaction Model

```{r}
uncSIM_Poisson = glm(formula = TRIPS ~ 
               log(ORIGIN_AGE7_12) +
               log(ORIGIN_AGE13_24) +
               log(ORIGIN_AGE25_64) +
               log(DESTIN_AGE7_12) +
               log(DESTIN_AGE13_24) +
               log(DESTIN_AGE25_64) +
               log(SCHOOL_COUNT) +
               log(BIZ_COUNT) + 
               log(DIST),
             family = poisson(link = "log"),
             data = inter_zonal_flow,
             na.action = na.exclude)

summary(uncSIM_Poisson)
```

### Origin (Production) Constrained Spatial Interaction Model

xxx

```{r}
orcSIM_Poisson = glm(formula = TRIPS ~
                       ORIGIN_SZ +
                       log(SCHOOL_COUNT) +
                       log(BIZ_COUNT) +
                       log(DIST) -1, # no need for intercept for origin/dest constrained
                     family = poisson(link = "log"),
                     data = inter_zonal_flow,
                     na.action = na.exclude)

options(max.print=10000) 
summary(orcSIM_Poisson)
```

```{r}
CalcRSquared = function(observed, estimated){
  r = cor(observed, estimated)
  R2 = r^2
  R2
}
```

The calculated R-squared value shows how well the factors explain the flow.

```{r}
CalcRSquared(orcSIM_Poisson$data$TRIPS, orcSIM_Poisson$fitted.values)
```

The RMSE (root mean square error) shows how much error the model typically makes in its predictions, with a higher weight for large errors. 

```{r}
performance_rmse(orcSIM_Poisson,
                 normalized = FALSE) # use raw values (not normalised to mean = 0, sd = 1)
```

xxx

```{r}
performance_rmse(orcSIM_Poisson,
                 normalized = TRUE)
```

### Destination Constrained Spatial Interaction Model

xxx

```{r}
decSIM_Poisson = glm(formula = TRIPS ~
                       DESTIN_SZ +
                       log(SCHOOL_COUNT) +
                       log(BIZ_COUNT) +
                       log(DIST) -1, # no need for intercept for origin/dest constrained
                     family = poisson(link = "log"),
                     data = inter_zonal_flow,
                     na.action = na.exclude)

options(max.print=10000) 
summary(decSIM_Poisson)
```

### Doubly Constrained Spatial Interaction Model

```{r}
dbcSIM_Poisson = glm(formula = TRIPS ~
                       ORIGIN_SZ +
                       DESTIN_SZ +
                       log(DIST), # note: -1 not required for doubly constrained SIM
                     family = poisson(link = "log"),
                     data = inter_zonal_flow,
                     na.action = na.exclude)
options(max.print=10000) 
summary(dbcSIM_Poisson)
```

### Model Comparison

```{r}
model_list = list(unconstrained = uncSIM_Poisson,
                  originConstrained = orcSIM_Poisson,
                  destinationConstrained = decSIM_Poisson,
                  doublyconstrained = dbcSIM_Poisson)
```

```{r}
compare_performance(model_list,
                    metrics = "RMSE")
```

The print above reveals that doubly constrained SIM is the best model among all the four SIMs because it has the smallest RMSE value of 1906.694.

[**\~\~\~ End of In-class Exercise 4 \~\~\~**]{.smallcaps}
