---
title: "Hands-on Exercise 1B"
date: "17 November 2023"
format: html
execute: 
  echo: true #shows the code
  eval: true #shows the outcomes of the code
  warning: false #does not show the warnings
editor: visual
background-color: lightgrey;
font-family:  Palatino Linotype;
---

# 2 Choropleth Mapping with R

## 2.1 Overview

In this page, I show how I had completed the Hands-on Exercise 1B, on plotting choropleth maps using the **tmap** package. It includes the codes used, the graphs obtained as well as the insights gathered.

Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors.

## 2.2 Getting Started

The R packages used in this hands-on exercises were:

-   **tmap** for thematic mapping;

-   **sf** for importing, managing, and processing geospatial data, and

-   **tidyverse** (i.e. readr, tidyr, dplyr) for performing data science tasks such as importing, tidying, and wrangling data.

They were loaded into the R environment using the following code:

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

***Student Note***: This allowed for loading of multiple packages in one line of code.

## 2.3 Importing Data

The following data sets were used in this hands-on exercise:

-   [Master Plan 2014 Subzone Boundary (Web)](https://beta.data.gov.sg/collections/1708/view) from [data.gov.sg](https://data.gov.sg/)

-   [Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020](https://www.singstat.gov.sg/find-data/search-by-theme/population/geographic-distribution/latest-data) from the [Department of Statistics Singapore](https://www.singstat.gov.sg/).

The data sets were placed under two sub-folders:

-   geospatial (Master Plan Subzone Boundary), and

-   aspatial (Residents).

These two sub-folders are within the data folder of my Hands-on_Ex1B folder.

### 2.3.1 Importing Geospatial Data

The **st_read()** (under sf package) was used to import the geospatial data set: `MP14_SUBZONE_WEB_PL`, a polygon feature layer in ESRI shapefile format.

```{r}
mpsz = st_read(dsn = "data/geospatial", 
                layer = "MP14_SUBZONE_WEB_PL")
```

***Student Note***: The geospatial objects are multipolygon features. There are a total of **323 features and 15 fields** in `mpsz` simple feature data frame. `mpsz` is in **svy21 projected coordinate system**. The bounding box provides the x extend and y extend of the data.

### 2.3.2 Importing Attribute Data

The **read_csv()** (under readr package) was used to import the aspatial data set: `respopagsex2000to2018`, a csv file with PA and SZ fields that can be used as unique identifiers to geocode to the `MP14_SUBZONE_WEB_PL` shapefile.

```{r}
popdata = read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### 2.3.3 Data Preparation

A data table was prepared to include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0-4 to age group 20-24;

-   ECONOMY ACTIVE: age group 25-29 to age group 60-64;

-   AGED: age group 65 and above;

-   TOTAL: all age groups; and

-   DEPENDENCY: the ratio between young + aged over economy active group.

The functions used were:

-   **pivot_wider()** (under tidyr package); and

-   **mutate()**, **filter()**, **group_by()** and **select()** (under **dplyr** package).

```{r}
popdata2020 = popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
  mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
           rowSums(.[13:15]))%>%
  mutate(`AGED`=rowSums(.[16:21])) %>%
  mutate(`TOTAL`=rowSums(.[3:21])) %>%  
  mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
         /`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
         `ECONOMY ACTIVE`, `AGED`, 
         `TOTAL`, `DEPENDENCY`)

popdata2020 = popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
            .funs = funs(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

The geographical data and attribute table were then joined using **left_join()** (under dplyr package).

```{r}
mpsz_pop2020 = left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))

write_rds(mpsz_pop2020, "data/mpszpop2020.rds")
```

## 2.4 Choropleth Mapping Geospatial Data Using tmap

### 2.4.1 Plotting A Choropleth Map Using qtm()

A cartographic standard choropleth map was drawn.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

### 2.4.2 Creating A Choropleth Map Using tmap's Elements

The **tmap** package is used to draw a high quality cartographic choropleth map.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

The step by step illustration of the basic building blocks of tmap are shown below:

Using **tm_shape()** and **tm_polygons()**:

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

Adding DEPENDENCY as target variable:

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

Using **tm_fill()** only:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

Using **tm_fill()** and **tm_borders()**:

***Student Note***: The combined use of **tm_fill()** and **tm_borders()** is the equivalent of using **tm_polygons()**.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

Using the style argument of **tm_fill()** or **tm_polygons()**:

```{r}

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

### 2.4.3 \[*Extra!*\] Choropleth Maps Using Different Classification Methods

Using different **styles**:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "pretty") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "hclust") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "bclust") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "log10_pretty") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "pretty") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fisher") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "dpih") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "headtails") +
  tm_borders(alpha = 0.5)
```

Using different number of **bins**:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "jenks") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "jenks") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "jenks") +
  tm_borders(alpha = 0.5)

tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "jenks") +
  tm_borders(alpha = 0.5)

```

Using **custom breaks**:

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

### 2.4.4 Colour Scheme

Using **ColourBrewer** palette:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

### 2.4.5 Map Layouts

Including **map legend**:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

Changing **map style**:

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

### 2.4.6 Facet Maps

Facet maps, are composed of many maps arrange side-by-side, and sometimes stacked vertically. Small multiple maps enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in three ways:

-   by assigning multiple values to at least one of the asthetic arguments,

-   by defining a group-by variable in **tm_facets()**, and

-   by creating multiple stand-alone maps with **tmap_arrange()**.

[Maps of distribution of young and aged populations using tm_fill() and tmap_arrange()]{.underline}

***Student Note***: The Northeast region more of the young, and less of the aged. This reflects the demographies of new towns like Punggol and Sengkang.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

```{r}
youngmap = tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap = tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

[Maps of distribution of aged populations and dependency ratios]{.underline}

***Student Note***: Even though the distribution of the aged varies across different subzones, the dependency levels are generally the same. The exception is Loyang West which has a very high dependency ratio.

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
              style = c("equal", "quantile"), 
              palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)

```

[**\~\~\~ End of Hands-on Exercise 1B \~\~\~**]{.smallcaps}
